{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or, How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-08-11 thesis/landslide.tex
describe estimation algs and make an example with RE (still tryna remember how WBE works tbh)

\subsection{Execution Tree}
\label{sec:landslide-save}

@@

\subsubsection{Time Travel}
\label{sec:landslide-timetravel}

@@

When should Landslide sunder the universe into alternate realities,
in which each a different thread executes immediately following the current instruction?
This singular question determines to what extent the state space of possible interleavings explodes exponentially.
While other parts of the research decide which lock API calls to consider,
or which memory accesses constitute a data race,
interpreting those combinations of preemption point {\em predicates}
to decide if the current program state constitutes a single preemption {\em point}
warrants discussion.

@@

Although a superhuman reader may quickly reach intuitive understanding
of complex algorithms from dense prose and mathematical notation alone,
I also present here a friendly example of using DPOR on Figure~\ref{fig:concurrency-bug}'s example program,
whose original state space is shown in Figure~\ref{fig:tree}.
%Acknowledging that quickly gaining an intuitive understanding for an algorithm
%is basically impossible from just paragraphs of text,
%I conclude this section with an example of a single iteration of DPOR
%on the example state space from Figures~\ref{fig:concurrency-bug} and \ref{fig:tree}.
First, let us consider a single iteration of DPOR, applied after executing the first branch.
The result is shown in Figure~\ref{fig:dpor-example-0}.

@@

\begin{figure}[h]
	\begin{center}
		\begin{tabular}{p{0.30\textwidth}p{0.30\textwidth}p{0.30\textwidth}}
	\begin{center}
	\begin{tabular}{l}
		\dporTAcode{tmp1 <- x} \\
		\dporTBcode{tmp2 <- x} \\
		\dporTAcode{x <- tmp1} \\
		\dporTBcode{x <- tmp2} \\
		% old example, pre dpor-example figure
		%$T_1$@$t_a$: \texttt{write~x} \\
		%$T_2$@$t_b$: \texttt{read~y} \\
		%$T_2$@$t_c$: \texttt{read~x} \\
	\end{tabular}
	\end{center}
	&
	\begin{center}
	\begin{tabular}{l}
		\dporTBcode{tmp2 <- x} \\
		\dporTBcode{x <- tmp2} \\
		\dporTAcode{tmp1 <- x} \\
		\dporTAcode{x <- tmp1} \\
		%$T_2$@$t_{b'}$: \texttt{read~y} \\
		%$T_2$@$t_{c'}$: \texttt{read~x} \\
		%$T_1$@$t_{a'}$: \texttt{write~x} \\
	\end{tabular}
	\end{center}
	&
	\begin{center}
	\begin{tabular}{l}
		\dporTBcode{tmp2 <- x} \\
		\dporTAcode{tmp1 <- x} \\
		\dporTAcode{x <- tmp1} \\
		\dporTBcode{x <- tmp2} \\
		%$T_2$@$t_{b'}$: \texttt{read~y} \\
		%$T_1$@$t_{a''}$: \texttt{write~x} \\
		%$T_2$@$t_{c''}$: \texttt{read~x} \\
	\end{tabular}
	\end{center}
	\\
	\begin{center}
	(a) Original branch (C).
	\end{center}
	&
	\begin{center}
	(b) Goal branch (K).
	\end{center}
	&
	\begin{center}
	(c) Redundant branch (L).
	\end{center}
	\end{tabular}
	\end{center}
	\caption{Motivating example for the sleep sets optimization,
	highlighting three interleavings from Figure~\ref{fig:dpor-example-2}.
	The always-independent {\tt tmpN++}s are omitted for brevity.
	}
	\label{fig:sleepsets}
\end{figure}

@@

Landslide implements this check in {\tt equiv\_already\_explored()},
which checks (in this case after executing K),
that if the first event to be reordered (here, \dporTA{{\tt tmp1 <- x}})
has already been tested in an equivalent reordering around any number of preceding events
(here, \dporTB{{\tt tmp2 <- x}} and \dporTB{{\tt tmp2++}}),
then the newly marked subtree is safe to prune.
This corresponds to the {\em sleep sets} optimization described by prior work \cite{partial-order-methods,dpor,optimal-dpor},
so named because it effectively puts \dporTA{$\mathbf{T_1}$}
``to sleep'' until after the true conflicting access of \dporTB{{\tt x <- tmp2}}.

Landslide's implementation differs from prior work,
which explicitly tracks sets of reordered threads and expected conflicting accesses,
by instead identifying
where the reduction should occur
during subsequent DPOR iterations.
% KEEP_RUNNING_DPORS_CHOSEN_TID
This approach also relies on the search ordering strategy ({\tt arbiter\_choose()})
to prefer scheduling the thread previously chosen for reordering by DPOR,
to ensure the conflicting access happens before the preempted thread gets a chance to run again.
%to force whatever its conflicting memory access was to happen before the other, preempted thread gets a chance to run.
Further optimizations such as {\em source sets} and {\em wakeup trees},
which prior work has shown achieve optimality
(i.e., executing exactly one interleaving per equivalence class) \cite{optimal-dpor}
are not yet implemented.
To the best of my knowledge,
they provide further reduction only in cases of 3 or more threads;
I suspect (without proof) that sleep-set DPOR is optimal for 2 threads.

@@

% TODO: once dpor example tree is made
% switch this to that
As a concrete example, consider the state space of Figure~\ref{fig:tree}(b) (untouched by DPOR),
and suppose each transition to the next preemption point takes 1 second to execute.
While the first branch executes in 5 seconds,
the second branch, sharing the first transition as a common prefix, takes 4 seconds,
and the one after that only 2;
the state space being ultimately completed in 33 seconds.
Even with perfect hindsight,
na\"ively multiplying the total interleavings (10)
by the total execution time per branch (5) would double-count common prefixes
and grossly overestimate (50) the total runtime.

Hence, Landslide uses two differently suitable algorithms for each of size and runtime estimation:
the Weighted Backtrack Estimator (WBE) and the Recursive Estimator (RE), respectively,
first introduced in \cite{estimating-search-tree-size} and later adapted to DPOR by \cite{estimation}.
In principle, both calculate the current progress as a proportion of the expected total
by counting how many branches DPOR has marked for future exploration (\sect{\ref{sec:landslide-explore}})
and assuming the sizes of their resulting subtrees are predicted by the known sizes of similar already-explored subtrees.
In practice, the calculation strategy differs between the two approaches,
which can occasionally result in drastically differing outputs (\sect{\ref{sec:tm-verif}}).
I summarize their respective implementations here.
% TODO more specific section reference
Chapter~\ref{chap:warpzone} will discuss their limitations and some opportunities for future improvement.

@@

@@

>>>>

\subsection{Execution tree}
\label{sec:landslide-save}

@@

\subsubsection{Time travel}
\label{sec:landslide-timetravel}

@@

When should Landslide sunder the universe into alternate realities,
in which each a different thread executes immediately following the current instruction?
This singular question determines to what extent the state space of possible interleavings explodes exponentially.
While other parts of
%the research
the great work % :relaxed:
decide which lock API calls to consider,
or which memory accesses constitute a data race,
interpreting those combinations of preemption point {\em predicates}
to decide if the current program state constitutes a single preemption {\em point}
warrants discussion.

@@

Although a superhuman reader may quickly reach intuitive understanding
of complex algorithms from dense prose and mathematical notation alone,
I also present here a friendly example of using DPOR on Figure~\ref{fig:concurrency-bug}'s example program,
whose original state space is shown in Figure~\ref{fig:tree}.
In this program both threads are always enabled, imposing no scheduling constraints,
so memory conflicts alone will drive exploration.
%Acknowledging that quickly gaining an intuitive understanding for an algorithm
%is basically impossible from just paragraphs of text,
%I conclude this section with an example of a single iteration of DPOR
%on the example state space from Figures~\ref{fig:concurrency-bug} and \ref{fig:tree}.
First, let us consider a single iteration of DPOR, applied after executing the first branch.
The result is shown in Figure~\ref{fig:dpor-example-0}.

@@

\begin{figure}[h]
	\begin{center}
		\begin{tabular}{p{0.30\textwidth}p{0.30\textwidth}p{0.30\textwidth}}
	\begin{center}
	\begin{tabular}{l}
		\dporTAcode{tmp1 <- x (read)} \\
		\dporTBcode{tmp2 <- x (read)} \\
		\dporTAcode{x <- tmp1 (write)} \\
		\dporTBcode{x <- tmp2 (write)} \\
		% old example, pre dpor-example figure
		%$T_1$@$t_a$: \texttt{write~x} \\
		%$T_2$@$t_b$: \texttt{read~y} \\
		%$T_2$@$t_c$: \texttt{read~x} \\
	\end{tabular}
	\end{center}
	&
	\begin{center}
	\begin{tabular}{l}
		\dporTBcode{tmp2 <- x (read)} \\
		\dporTBcode{x <- tmp2 (write)} \\
		\dporTAcode{tmp1 <- x (read)} \\
		\dporTAcode{x <- tmp1 (write)} \\
		%$T_2$@$t_{b'}$: \texttt{read~y} \\
		%$T_2$@$t_{c'}$: \texttt{read~x} \\
		%$T_1$@$t_{a'}$: \texttt{write~x} \\
	\end{tabular}
	\end{center}
	&
	\begin{center}
	\begin{tabular}{l}
		\dporTBcode{tmp2 <- x (read)} \\
		\dporTAcode{tmp1 <- x (read)} \\
		\dporTAcode{x <- tmp1 (write)} \\
		\dporTBcode{x <- tmp2 (write)} \\
		%$T_2$@$t_{b'}$: \texttt{read~y} \\
		%$T_1$@$t_{a''}$: \texttt{write~x} \\
		%$T_2$@$t_{c''}$: \texttt{read~x} \\
	\end{tabular}
	\end{center}
	\\
	\begin{center}
	(a) Original branch (C).
	\end{center}
	&
	\begin{center}
	(b) Goal branch (K).
	\end{center}
	&
	\begin{center}
	(c) Redundant branch (L).
	\end{center}
	\end{tabular}
	\end{center}
	\caption{Motivating example for the sleep sets optimization.
	Three of Figure~\ref{fig:dpor-example-2}'s interleavings are highlighted,
	with the always-independent {\tt tmpN++}s omitted for brevity.
	}
	\label{fig:sleepsets}
\end{figure}

@@

Landslide implements this check in {\tt equiv\_already\_explored()},
which checks (in this case after executing K),
that if the first event to be reordered (here, \dporTA{{\tt tmp1 <- x}})
has already been tested in an equivalent reordering around any number of preceding events
(here, \dporTB{{\tt tmp2 <- x}} and \dporTB{{\tt tmp2++}}),
then the newly marked subtree is safe to prune.
Note that this does not require storing any full subtrees outside of the current branch;
only the subtree's root node need be saved to
prove that an equivalent interleaving beginning with \dporTA{$\mathbf{T_1}$} therein was already checked,
preserving DPOR's $O(n)$ memory footprint.

This corresponds to the {\em sleep sets} optimization described by prior work \cite{partial-order-methods,dpor,optimal-dpor},
so named because it effectively puts \dporTA{$\mathbf{T_1}$}
``to sleep'' until after the true conflicting access of \dporTB{{\tt x <- tmp2}}.
Landslide's implementation differs from prior work,
which explicitly tracks sets of reordered threads and expected conflicting accesses,
by instead identifying
where the reduction should occur
during subsequent DPOR iterations.
% KEEP_RUNNING_DPORS_CHOSEN_TID
This approach also relies on the search ordering strategy ({\tt arbiter\_choose()})
to prefer scheduling the thread previously chosen for reordering by DPOR,
to ensure the conflicting access happens before the preempted thread gets a chance to run again.
%to force whatever its conflicting memory access was to happen before the other, preempted thread gets a chance to run.
Further optimizations such as {\em source sets} and {\em wakeup trees},
which prior work has shown achieve optimality
(i.e., executing exactly one interleaving per equivalence class) \cite{optimal-dpor}
are not yet implemented.
To the best of my knowledge,
they provide further reduction only in cases of 3 or more threads;
I suspect (without proof) that sleep-set DPOR is optimal for 2 threads.

@@

As a concrete example, consider the state space of Figure~\ref{fig:dpor-example-2},
and suppose each transition to the next preemption point takes 1 second to execute.
While the first branch executes in 6 seconds,
the second branch, sharing the first transition as a common prefix, takes 4 seconds,
and the one after that only 2;
the state space being ultimately completed in 24 seconds.
Even with perfect hindsight,
na\"ively multiplying the total interleavings (6)
by the total execution time per branch (6) would double-count common prefixes
and grossly overestimate (36) the total runtime.

Hence, Landslide uses two differently suitable algorithms for each of size and runtime estimation:
the Weighted Backtrack Estimator (WBE) and the Recursive Estimator (RE), respectively,
first introduced in \cite{estimating-search-tree-size} and later adapted to DPOR by \cite{estimation}.
In principle, both calculate the current progress as a proportion of the expected total
by counting how many branches DPOR has marked for future exploration (\sect{\ref{sec:landslide-explore}})
and assuming the sizes of their resulting subtrees are predicted by the known sizes of similar already-explored subtrees.
In practice, the calculation strategy differs between the two approaches,
which can occasionally result in drastically differing outputs (\sect{\ref{sec:tm-verif}}).

Implementation-wise, Landslide reports size estimates as both the percentage and as a total number of branches,
and time estimates as an ETA.
Quicksand's {\tt -v} option (\sect{\ref{sec:landslide-quicksand-options}})
will cause it to print them each time a new interleaving is tested; for example:
\begin{center}
	{\tt \small [JOB 1] progress: 66101/94825 brs (69.708252\%), ETA 13m 37s (elapsed 46m 10s)}
\end{center}
Both estimates are computed simultaneously in {\tt \_estimate()} in {\tt estimate.c}
(which, I might add, is well-commented in case the following prose is insufficient).
%I summarize their respective implementations here.
% TODO more specific section reference
Chapter~\ref{chap:warpzone} will discuss their limitations and some opportunities for future improvement.

@@

The WBE, used to estimate total number of interleavings,
computes the {\em proportion} of the total size
that the already-explored branches are expected to comprise,
using DPOR's workqueue to anticipate how many unexplored marked branches remain.
This serves as a ``progress bar'' that represents the estimated overall progress,
approaching 1 (not necessarily monotonically) as exploration continues.

Summarizing prior work's formal definition \cite{estimation},
the proportion at a terminal node $v_n$%
\footnote{Prior work \cite{estimating-search-tree-size,estimation} refers to this instead as {\em probability},
i.e., the probability that the node will appear in a branch chosen uniformly at random from the completed tree.
I find ``proportion'' to be more illuminating on how the algorithm works.
},
preceded by an execution sequence $(v_1 \dots v_{n-1})$,
is computed as:%
\footnote{
Simplified from \cite{estimation}: the missing $F(v_i)$ is 0, using the empty fit strategy.
}

\[
	\mathsf{proportion}(v_1 \dots v_n) = \displaystyle\prod_{i=1}^{n-1} \frac{1}{|\mathsf{marked~children}(v_i)|}
\]

where $\mathsf{marked~children}(v_i)$ is the number of enabled thread transitions at $v_i$
which have either already been explored or been marked by DPOR.
Then, the total estimate is given as the sum over all branches $b = (v_1 \dots v_n)$ explored so far:%
\footnote{
Simplified from \cite{estimation}: $t(b)$, the time for each branch, is 1, because we are counting them.
}

\[
	\mathsf{estimate} = \frac{1}{\sum_{b \in B} \mathsf{proportion}(b)}
\]

It is easy to see how these formulas might fit into DPOR's incremental search procedure:
at the end of each branch compute its $\mathsf{proportion}$ and add it in to a global $\mathsf{estimate}$ value.
However, DPOR may tag new branches to explore that would affect past branches' proportions,
in which case they would have to be recomputed,
which is not feasible without storing the entire exponentially-sized tree in memory.
Instead, Landslide also stores per-subtree estimates at intermediate $v_i$ nodes, $1 < i < n$,
along the current branch.
Whenever DPOR marks a new $k$th branch for exploration
at some $v_i$
%with $\mathsf{marked~children}(v_i) = k$,
its estimate
is multiplied by $(k-1)/k$ to retroactively adjust all past branches' proportions contributing to that estimate.
The change is also propagated to its sub-subtrees,
whose estimates must also incorporate the new $\mathsf{marked~children}$ value.
This allows Landslide to update the global estimate after each branch in $O(n)$ time and memory,
without recomputing past branch proportions individually.

@@

The RE, used to estimate total execution time,
computes at each node the expected time to execute all subtrees rooted at children of that node,
assuming unvisited subtrees' times will be an average of their visited siblings.
This estimate at the root node, minus the current time elapsed so far,
serves as a guess at how long until completion.
Let $\mathsf{usecs}(v_i)$ denote the time elapsed during execution of the transition $v_{i-1} \rightarrow v_i$.
Then a node's estimate is given by:

\[
	\mathsf{estimate}(v_i) = \mathsf{usecs}(v_i) +
	\frac{|\mathsf{marked~children}(v_i)|}{|\mathsf{explored~children}(v_i)|}
	\sum_{v_j \in \mathsf{explored~children}(v_i)} \mathsf{estimate}(v_j)
\]

Like the WBE, whenever DPOR tags a new $k$th child at some $v_i$,
its estimate is multiplied by $k/(k-1)$ (note the reciprocal of before)
to retroactively re-weight previously explored subtrees' estimates.
Unlike the WBE, this change does not need to be propagated to descendant subtrees' estimates.
This estimate also takes $O(n)$ time and memory.

\subsubsection{Accuracy example}

To illustrate how the two estimators can under-estimate the total tree size and/or diverge from each other,
consider the state space from Figure~\ref{fig:dpor-example-2}, of size 6.
Suppose for RE that each transition takes 1 second to execute.

\begin{enumerate}
	\item After branch A, two tags exist, $\bigstar$1 and $\bigstar$3.

		Under RE, the estimate at \dporTA{{\tt tmp1++}} will be 9 seconds,
		and the root estimate will be 20 seconds, an underestimate.
	\item After branch B, $\bigstar$2 is now marked.

		Under RE, the estimate at \dporTB{{\tt tmp2++}} is 5 seconds,
		the estimate at \dporTA{{\tt tmp1++}} is updated to 11 seconds,
		and the root estimate to 24 seconds, accurate.
	\item After branch C, nothing new was marked.

		Under RE, no estimates change from after B.
	\item After branch K, $\bigstar$4 now exists.

		Under RE, the estimate at \dporTB{{\tt tmp2++}} is 9 seconds,
		and the root estimate is 22 seconds, an underestimate.
	\item After branch L, $\bigstar$5 joins the party.

		Under RE, the estimate at \dporTB{{\tt tmp2++}} is 11 seconds,
		and the root estimate is 24 seconds, accurate.
	\item After branch M, both estimators have perfect information and converge to accuracy.
\end{enumerate}

To illustrate how the estimators can over-estimate the total tree size,
consider the same state space,
except with the $\bigstar$4 subtree also pruned by DPOR's sleep sets extension (\sect{\ref{sec:landslide-sleepsets}});
i.e., only branches A, B, C, and K remain,
with an 18 second execution time.
Both estimators' behaviour is identical through branch C,
only now WBE's prediction is accurate at A but overestimates at B and C,
while RE's predictions are all overestimates.
As before, both reach perfect accuracy upon completion, now occurring at K.
Intuitively speaking, the estimators underpredict when DPOR keeps finding new branches to tag as it makes progress,
and overpredict when sleep set reduction achieves extra pruning on right subtrees.
Not shown in this example, interleaving-dependent control flow can, of course,
beget unexpected state space structure in essentially arbitrary other ways.
